Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-21 analysis/paper/paper.Rmd
add in-text code

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007]. Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.

@@

A total of 73 vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom. The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.

## metric measurements

@@

phases_isect <- 
  pottery_data %>%  
  select(filename, phase) %>% 
  filter(filename %in% isect) %>% 
  mutate_all(as.factor) %>% 
  distinct(filename, phase) %>% 
  mutate(phase = as.factor(case_when(
    phase == 'post-e' ~ "Post-European",
    phase == 'pre-e' ~ "Pre-European",
    phase == 'ch-con' ~ "Chinese" # we combine Chinese and post-Euro
  )))

# how many pots in each phase?
phases_isect %>% 
  group_by(phase) %>% 
  tally(sort = T)
```

@@

The elliptic Fourier coefficients of 73 pottery from three different phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.

@@

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) can explain `r vari_two`% of variances, and the first three principal components (PCs) can explain `r vari_three`% of variances. The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight. The biplot of PCs presents the grouping by each occupation phase.

>>>>

Geometric morphometric methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring length, width or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates by their unique point location with respect to numerical values of coordinate axes. According to different focuses and nature of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks or semilandmarks onto objects as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions, or x, y, z coordinates for three dimensions. Based on distances between raw landmarks, objects are translated to a common centroid, rescaled into the same size, and rotated until the summed squared distances between corresponding landmarks are minimized using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA is a method that superimposes sets of landmark configurations into a common coordinate system, where superimposed landmark coordinates are used as shape variables for further multivariate statistics analyses[@Slice2007]. Landmark-based morphometrics are widely used for archaeological objects where morphological signature is obvious to identity which provides ideal reference points for landmark placement, such as projectile point tips or biologically measurements points indicating osteological features [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for complex or rounded shaped structures. In those cases, outline approaches serve as a solution by analyzing the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, also called sliding landmarks, which assigns points along the curve between two landmarks at defined intervals [@Bookstein1997; @Lawing2010]. Those semi-landmarks are allowed to slide along the curve to remove the effect of the arbitrary landmark spacing by minimizing either Procrustes distance or bending energy [@Bookstein1997; @Slice2007; @Gunz2013]. Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adams2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.

@@

phases_isect <- 
  pottery_data %>%  
  select(filename, phase) %>% 
  filter(filename %in% isect) %>% 
  mutate_all(as.factor) %>% 
  distinct(filename, phase) %>% 
  mutate(phase = as.factor(case_when(
    phase == 'post-e' ~ "Post-European", # include European period and after it before Chinese contact
    phase == 'pre-e' ~ "Pre-European",
    phase == 'ch-con' ~ "Chinese" # we can combine Chinese and post-Euro
  )))

# how many pots in each phase?
phase_fre <- phases_isect %>% 
  group_by(phase) %>% 
  tally(sort = T)

total_no <- sum(phase_fre$n)
```

A total of `r total_no` vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for this study because of their completeness mostly covering parts from rim to bottom. The scanned pottery drawings of those pots were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline.

## Metric measurements

@@

@@

The elliptic Fourier coefficients of `r total_no` pottery from three different phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.

@@

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances. The PC1 relate to the height of the vessels from short to high, the PC2 reflects the width of their necks and mouths from narrow to wide, and the PC3 relates to the degree of flare of neck from curve to straight. The biplot of PCs presents the grouping by each occupation phase.
